---
# Development knitting only - ignored by Quarto book build
output:
  word_document:
    reference_docx: templates/custom-reference.docx
bibliography: bib/references.bib
csl: csl/csas.csl
link-citations: true
---

## Understanding Salmon and Their Ecosystems {#sec:understanding-salmon-ecosystems}

```{r setup, include=FALSE}
source(here::here("_common.R"))
```

Context for this section

### A. Salmon population monitoring

```{r salmon-population-projects}
library(flextable)

con <- get_db_connection()
s_monitor_projects <- dbGetQuery(con, "
  SELECT project_id, source, project_title
  FROM projects
  WHERE include = 'y'
    AND broad_header = 'Understanding salmon ecosystems'
    AND section = 'Salmon population monitoring'
  ORDER BY project_id
")
dbDisconnect(con)

flextable(s_monitor_projects) %>%
  set_header_labels(
    project_id = "Project",
    source = "Source",
    project_title = "Title"
  ) %>%
  width(j = 1, width = 1) %>%
  width(j = 2, width = 1) %>%
  width(j = 3, width = 4.5) %>%
  set_caption("Projects related to salmon population monitoring") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  bold(part = "header") %>%
  align(align = "left", part = "all") %>%
  valign(valign = "top", part = "body")
```

### B. Life cycle approaches

#### Follow the Fish (IP)

Provide an overview of the follow the fish program, how risk assessments and limiting factors (RAMS) were used to guide research priorities, key insights, and next steps to support rebuilding of WCVI Chinook salmon.

```{r ftf-projects}
library(flextable)

con <- get_db_connection()
ftf_projects <- dbGetQuery(con, "
  SELECT project_id, source, project_title
  FROM projects
  WHERE include = 'y'
    AND broad_header = 'Assessments'
    AND section = 'Life cycle assessment - Follow The Fish'
  ORDER BY project_id
")
dbDisconnect(con)

flextable(ftf_projects) %>%
  set_header_labels(
    project_id = "Project",
    source = "Source",
    project_title = "Title"
  ) %>%
  width(j = 1, width = 1) %>%
  width(j = 2, width = 1) %>%
  width(j = 3, width = 4.5) %>%
  set_caption("Projects related to life cycle assessment - Follow The Fish") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  bold(part = "header") %>%
  align(align = "left", part = "all") %>%
  valign(valign = "top", part = "body")
```

#### Bottlenecks to survival (IP)

```{r bottleneck-projects}
library(flextable)

con <- get_db_connection()
bottleneck_projects <- dbGetQuery(con, "
  SELECT project_id, source, project_title
  FROM projects
  WHERE include = 'y'
    AND broad_header = 'Assessments'
    AND section = 'Life cycle assessment - Bottlenecks to survival'
  ORDER BY project_id
")
dbDisconnect(con)

flextable(bottleneck_projects) %>%
  set_header_labels(
    project_id = "Project",
    source = "Source",
    project_title = "Title"
  ) %>%
  width(j = 1, width = 1) %>%
  width(j = 2, width = 1) %>%
  width(j = 3, width = 4.5) %>%
  set_caption("Projects related to life cycle assessment - Bottlenecks to survival") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  bold(part = "header") %>%
  align(align = "left", part = "all") %>%
  valign(valign = "top", part = "body")
```

### C. Freshwater ecosystems (JE)

```{r freshwater-ecosystem-projects}
library(flextable)

con <- get_db_connection()
freshwater_projects <- dbGetQuery(con, "
  SELECT project_id, source, project_title
  FROM projects
  WHERE include = 'y'
    AND broad_header = 'Understanding salmon ecosystems'
    AND section = 'Freshwater ecosystems'
  ORDER BY project_id
")
dbDisconnect(con)

flextable(freshwater_projects) %>%
  set_header_labels(
    project_id = "Project",
    source = "Source",
    project_title = "Title"
  ) %>%
  width(j = 1, width = 1) %>%
  width(j = 2, width = 1) %>%
  width(j = 3, width = 4.5) %>%
  set_caption("Projects related to freshwater ecosystems") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  bold(part = "header") %>%
  align(align = "left", part = "all") %>%
  valign(valign = "top", part = "body")
```

Summarize research undertaken and key findings focused on freshwater ecosystems. Projects and themes include:

-   Risk assessment and mitigation of aquatic invasive species

    -   Understanding threats of freshwater AIS (Thomas Therriault)

    -   Survival and predation on Okanagan Chinook (Tommy Pontbriand)

    -   Monitoring of Smallmouth bass in Cultus Lake (Daniel Doutaz)

-   Research on freshwater toxins and their effects on salmon, particulalrly 6PPD-Q

    -   Identifying tire-related toxic hotspots (Lisa Loseto)

-   Physiological and mechanistic studies to link environmental conditions with survival and productivity

    -   Mechanistic modelling linking hydrology to habitat quality (Sean Naman)

    -   Physiological studies to understand cumulative impacts on salmon survival (David Patterson)

    -   Effects of redd scour on WCVI Chinook (Diana McHugh)

-   Broad-scale spatial and statistical modelling to characterize cumulative threats to freshwater habitat

    -   Geospatial indicators of cumulative threats to stream habitat (Josie Iacarella)

    -   Effects of forest harvest of salmon productivity (BCSRIF, Maria Kuruvilla)

    -   The CEMPRA cumulative effects modelling framework (BCSRIF, Jordan Rosenfeld)

### D. Marine ecosystems (IP/NB)

```{r marine-ecosystem-projects}
library(flextable)

con <- get_db_connection()
marine_projects <- dbGetQuery(con, "
  SELECT project_id, source, project_title
  FROM projects
  WHERE include = 'y'
    AND broad_header = 'Understanding salmon ecosystems'
    AND section = 'Marine ecosystems'
  ORDER BY project_id
")
dbDisconnect(con)

flextable(marine_projects) %>%
  set_header_labels(
    project_id = "Project",
    source = "Source",
    project_title = "Title"
  ) %>%
  width(j = 1, width = 1) %>%
  width(j = 2, width = 1) %>%
  width(j = 3, width = 4.5) %>%
  set_caption("Projects related to marine ecosystems") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  bold(part = "header") %>%
  align(align = "left", part = "all") %>%
  valign(valign = "top", part = "body")
```

Summarize research and key findings related to building understanding of marine ecosystems and key stressors. Projects and themes include:

-   Oceanographic studies monitoring marine conditions to understand their influence on salmon productivity

    -   Zooplankton and krill monitoring (Akash Sastri)

    -   Hypoxia and temperature within Vancouver Island fjords (Laura Bianucci)

    -   Acoustic monitoring of salmon and salmon prey (Stephane Gauthier)

    -   Monitoring harmful algal biotoxins (Andrew Ross)

    -   Contaminants monitoring (Lisa Loseto)

-   Research and monitoring on marine ecosystems and predators

    -   eDNA monitoring in WCVI (Christoph Deeg)

    -   Quantifying salmon predation by sea lions (Cameron Freshwater)

### E. Climate change vulnerability (CL)

```{r climate-projects}
library(flextable)

con <- get_db_connection()
climate_projects <- dbGetQuery(con, "
  SELECT project_id, source, project_title
  FROM projects
  WHERE include = 'y'
    AND broad_header = 'Understanding salmon ecosystems'
    AND section = 'Climate change vulnerabiliity'
  ORDER BY project_id
")
dbDisconnect(con)

flextable(climate_projects) %>%
  set_header_labels(
    project_id = "Project",
    source = "Source",
    project_title = "Title"
  ) %>%
  width(j = 1, width = 1) %>%
  width(j = 2, width = 1) %>%
  width(j = 3, width = 4.5) %>%
  set_caption("Projects related to climate change vulnerability") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  bold(part = "header") %>%
  align(align = "left", part = "all") %>%
  valign(valign = "top", part = "body")
```

**Highlights**

-   Newly developed marine and freshwater climate models allow us to assess environmental changes that will be experience by Pacific salmon within British Columbia

-   Vulnerability assessments reveal differences in risk for the 50 conservation units of Pacific salmon within the Fraser basin, with Pink and Chum salmon the least vulnerable to climate changes while ….

-   Life stage models….

-   Genetic adaptation….

-   Greenhouse gas emissions matter. Vulnerability, as represented by conditions beyond optimal thresholds, was more severe in high emissions scenarios compared to scenarios with mitigation.

-   These new assessments can be applied in planning and decision-making to contextualize risks and benefits….

**Research Goals**

The significance of climate change for Pacific salmon, and biodiversity in general, cannot be overstated. Studies across their range have highlighted the vulnerability of Pacific salmon to recent and future climate changes for freshwater and marine life stages [@mantua2010; @mcclure2023; @crozier2019; @grant2019]. Climate change emerged as a key theme during the PSSI Science planning workshop [@lagasse2024] and was a central element of the Improved Understanding of Salmon Ecosystems sub-initiative. Discussion papers, minister mandate letters, funding initiatives, and national adaptation plans have recognized the need to make climate change considerations central to decision-making and planning [@macdonald2023]. This need follows from the understanding that climate change introduces additional risk and uncertainty into plans and decisions because expectations based on past climate relationships may not hold true.

Yet there is still much that is uncertain about how climate changes will unfold for Pacific salmon, such as which species and stocks will be most affected, when physiological thresholds or ecological tipping points will be reached, the relative influence of various environmental changes on productivity, and the degree to which populations will be able to evolve and adapt to changing conditions and increasing temperatures [@schoen2017; @chittenden2009; @crozier2025]. Research within this theme sought to better understand and resolve these uncertainties through a variety of approaches including downscaled climate modelling, modern genotyping, life-cycle modelling, and vulnerability risk assessments. Collectively, these studies reveal important details about the implications of climate change for Pacific salmon that can allow us to understand and anticipate current and future changes.

**Downscaled climate modelling**

Global climate modelling – the simulation of earth systems into the future under different greenhouse gas emission scenarios – is the basis for much of our understanding about future climate change as reported on by the intergovernmental panel on climate change (IPCC). However, these global models lack the necessary detail to tease out fine-scale environmental changes to aquatic habitats that will be significant to understanding responses of Pacific salmon over the coming decades. Recent work, including many projects supported by PSSI, is addressing these gaps through downscaling of global climate models to marine and freshwater environments across the Pacific region.

Within freshwater habitats, the Pacific Climate Impacts Consortium (PCIC) has developed high resolution stream temperature and flow models that project future conditions throughout British Columbia. Statistically downscaled models of August stream temperature have also recently been developed providing an alternative method for projecting future temperatures.  These models reveal a gradient of temperature changes throughout British Columbia influenced not only by global climate patterns, but by elevation, glacial coverage, latitude…..

Within marine areas, regional oceanographic models developed by the Ocean Science division project changes in salinity, temperature, and biogeochemistry for waters across the continental shelf of British Columbia. While these models have recently been applied to understand how climate change will affect habitat availability for groundfish and other marine species, applying them towards the assessment of climate change vulnerability for Pacific salmon is more challenging…  Given the importance of marine survival to the productivity of Pacific salmon, this is an important area for further research….

**Vulnerability assessments**

For the purposes of Pacific salmon management, projecting future environmental conditions is only helpful if we can predict how Pacific salmon will respond to those changes. Climate change vulnerability assessments apply a diverse range of methods to consider how the combined exposure and sensitivity of a species or population determines the risks they face from climate change. While such assessments have been applied to thousands of different species at regional and global scales, including to Pacific salmon in the continental US, it is only recently and with support from PSSI that these studies have been applied to Pacific salmon in Canada.

Two of these studies have been focused on the Fraser Basin, the largest and most diverse river basin for Pacific salmon in Canada with over 50 conservation units. The Pacific Salmon Foundation developed a model to quantify life-stage specific exposure of Pacific salmon conservation units to environmental conditions above optimal thresholds, including high temperature and low flow thresholds. DFO Science has developed a climate vulnerability indicators framework that incorporates several downscaled climate models, demographic information such as conservation status, and genetic diversity information to characterize key mechanisms of vulnerability.

Together, these studies reveal …..

Life stage modelling can also be used for evaluating and disentangling the influence of environmental conditions on salmon productivity retroactively and into the future. Models of sockeye salmon, covering populations from the southern to northern ends of their range, reveal important similarities and differences in responses….

**Genetic adaptation**

**Knowledge gaps and future priorities**

Expanding vulnerability assessments to populations beyond the Fraser basin

Further understanding thresholds and tipping points, and how these differ among species of Pacific salmon

Changes in the frequency of extreme events, such as flooding, marine heat-waves, and droughts, are difficult to accurately model and assess but are important determinants of salmon survival

Incorporating climate vulnerability information into decision-making and planning frameworks, prioritizing actions that can mitigate vulnerability

\newpage
