## Science to support hatchery modernization (CL/SEP) {#sec:hatcheries}

```{r setup, include=FALSE}
source(here::here("_common.R"))
```

Describe the enhancement pillar of PSSI, DFO Science sub-initiatives, and the priority research areas. Include a description of the SEP-SCI steering committee and process used to prioritize and develop research projects (adapt elements from the SEP hatchery science booklet).

### A. Genetics

```{r genetics-projects}
library(flextable)

con <- get_db_connection()
genetics_projects <- dbGetQuery(con, "
  SELECT project_number, source, project_title
  FROM projects
  WHERE include = 'y'
    AND broad_header = 'Science to support hatchery modernization'
    AND section = 'Genetics'
  ORDER BY project_number
")
dbDisconnect(con)

flextable(genetics_projects) %>%
  set_header_labels(
    project_number = "Project",
    source = "Source",
    project_title = "Title"
  ) %>%
  width(j = 1, width = 1) %>%
  width(j = 2, width = 1) %>%
  width(j = 3, width = 4.5) %>%
  set_caption("Projects related to genetics") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  bold(part = "header") %>%
  align(align = "left", part = "all") %>%
  valign(valign = "top", part = "body")
```

Summarize research and key findings from genetics projects. Research questions (from SEP):

-   What are the differences between hatchery and natural origin fish?

-   Why do these differences occur and what are the mechanisms behind them?

-   What factors affect long-term population fitness?

Projects and themes include:

-   Developing and operationalizing modern genetic methods for stock identification and assessment

    -   Continuation/expansion of PBT programs (Eric Rondeau)

    -   Improving stock delineation in Chum salmon by characterizing genome-wide genetic variation (Anna Tigano)

    -   Rapid broodstock screening for improved genetic management of Pacific salmon (Kyle Wellband)

-   Evaluating genetic differences among hatchery and wild populations, and their implications for fitness

    -   Fitness effects of hatchery enhancement and relative reproductive success (Tim Healy)

    -   Epigenetic variation and inheritance in hatchery and natural origin Chinook and Coho salmon (Tim Healy, Kyle Wellband)

    -   Genetic variation linked to early maturation in male Chinook salmon

### B. Fish health and hatchery practices

```{r health-projects}
library(flextable)

con <- get_db_connection()
health_projects <- dbGetQuery(con, "
  SELECT project_number, source, project_title
  FROM projects
  WHERE include = 'y'
    AND broad_header = 'Science to support hatchery modernization'
    AND section = 'Fish health and hatchery practices'
  ORDER BY project_number
")
dbDisconnect(con)

flextable(health_projects) %>%
  set_header_labels(
    project_number = "Project",
    source = "Source",
    project_title = "Title"
  ) %>%
  width(j = 1, width = 1) %>%
  width(j = 2, width = 1) %>%
  width(j = 3, width = 4.5) %>%
  set_caption("Projects related to fish health and hatchery practices") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  bold(part = "header") %>%
  align(align = "left", part = "all") %>%
  valign(valign = "top", part = "body")
```

Summarize research and key findings from fish health projects. Research questions (from SEP):

-   How can we better assess and track fish condition and stress?

-   How do we advance aFish health and hatchery best practicesnd modernize diagnostic and health-management practices?

-   How can new molecular and technological tools help us detect issues early and inform decision-making?

Projects and themes include:

-   New diagnostic and monitoring tools for fish health

    -   Expanding and modernizing the finfish diagnostic lab (Amy Long)

    -   Developing a proactive, modernized approach to ensure optimal health and condition (Karia Kaukinen, Kristi Miller)

    -   Using scales to monitor chronic stress in Chinook and Coho salmon (Stewart Johnson)

    -   Application of nanopore technology for detection and characterization of pathogens (BCSRIF, Ahmed Siah)

-   Evaluating factors determining fish health and disease prevalence

    -   Using broodstock data to understand factors that affect BKD rates (Amy Long)

    -   Factors influencing IHNV in sockeye salmon (Kyle Garver)

    -   Modeling interactions of environmental, biological, and infectious factors at the Quinsam river hatchery (Simon Jones)

-   Effects of rearing practices and diet on hatchery fish survival

    -   Optimizing feeds for hatchery production of Pacific salmon (Erika Eliason)

    -   Predicting Chinook reproductive success based on adult thiamine levels (Erika Eliason)

    -   Causes and consequences of vateritic otoliths in hatchery-reared Coho salmon (BCSRIF, Leigh Gaffney)
    
\newpage