---
# Development knitting only - ignored by Quarto book build
output:
  word_document:
    reference_docx: templates/custom-reference.docx
bibliography: bib/references.bib
csl: csl/csas.csl
link-citations: true
---

## Restoration and recovery {#sec:restoration-recovery}

```{r setup, include=FALSE}
source(here::here("_common.R"))
```

### A. Recovery and rebuilding Programs

```{r recovery-projects}
library(flextable)

con <- get_db_connection()
recovery_projects <- dbGetQuery(con, "
  SELECT project_id, source, project_title
  FROM projects
  WHERE include = 'y'
    AND broad_header = 'Restoration and recovery'
    AND section = 'Recovery and rebuilding programs'
  ORDER BY project_id
")
dbDisconnect(con)

flextable(recovery_projects) %>%
  set_header_labels(
    project_id = "Project",
    source = "Source",
    project_title = "Title"
  ) %>%
  width(j = 1, width = 1) %>%
  width(j = 2, width = 1) %>%
  width(j = 3, width = 4.5) %>%
  set_caption("Projects related to recovery and rebuilding programs") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  bold(part = "header") %>%
  align(align = "left", part = "all") %>%
  valign(valign = "top", part = "body")
```

### B. Restoration prioritization

```{r restoration-projects}
library(flextable)

con <- get_db_connection()
restoration_projects <- dbGetQuery(con, "
  SELECT project_id, source, project_title
  FROM projects
  WHERE include = 'y'
    AND broad_header = 'Restoration and recovery'
    AND section = 'Restoration prioritization'
  ORDER BY project_id
")
dbDisconnect(con)

flextable(restoration_projects) %>%
  set_header_labels(
    project_id = "Project",
    source = "Source",
    project_title = "Title"
  ) %>%
  width(j = 1, width = 1) %>%
  width(j = 2, width = 1) %>%
  width(j = 3, width = 4.5) %>%
  set_caption("Projects related to restoration prioritization") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  bold(part = "header") %>%
  align(align = "left", part = "all") %>%
  valign(valign = "top", part = "body")
```

### Template / Guiding Questions

Research Goals

-   What were the key objectives of this research theme?

-   How do these relate to original goals of PSSI/BCSRIF?

-   What major gaps in our knowledge were being addressed?

Outcomes

-   What have we learned and how have we moved the dial on our scientific understanding?

-   What new tools/tech/techniques have been created?

-   How can these projects inform management planning/actions/policy?

Engagement, Collaboration & Knowledge Transfer

-   How did these projects engage with other groups? Was this effective? 

-   Have new ways to ensure knowledge transfer to different audiences been developed?

-   What products have been created? 

Next Steps

-   Major conclusions

-   How could/should the project findings be operationalized for salmon conservation and management?

-   Remaining knowledge gaps and recommendations for future studies

\newpage