---
# Development knitting only - ignored by Quarto book build
output:
  word_document:
    reference_docx: templates/custom-reference.docx
bibliography: bib/references.bib
csl: csl/csas.csl
link-citations: true
---

## Restoration and recovery {#sec:restoration-recovery}

```{r setup, include=FALSE}
source(here::here("_common.R"))
```

### A. Recovery and rebuilding Programs

```{r recovery-projects}

make_section_table(projects_df,
                   section_pick = 'Recovery and rebuilding programs')


```

### B. Restoration prioritization

```{r restoration-projects}

make_section_table(projects_df,
                   section_pick = 'Restoration and prioritization')


```

### Template / Guiding Questions

Research Goals

-   What were the key objectives of this research theme?

-   How do these relate to original goals of PSSI/BCSRIF?

-   What major gaps in our knowledge were being addressed?

Outcomes

-   What have we learned and how have we moved the dial on our scientific understanding?

-   What new tools/tech/techniques have been created?

-   How can these projects inform management planning/actions/policy?

Engagement, Collaboration & Knowledge Transfer

-   How did these projects engage with other groups? Was this effective? 

-   Have new ways to ensure knowledge transfer to different audiences been developed?

-   What products have been created? 

Next Steps

-   Major conclusions

-   How could/should the project findings be operationalized for salmon conservation and management?

-   Remaining knowledge gaps and recommendations for future studies

\newpage