## Restoration and recovery {##sec:restoration-recovery}

### A. Recovery and rebuilding Programs

```{r recovery-projects}
con <- get_db_connection()
recovery_projects <- dbGetQuery(con, "
  SELECT project_number, source, project_title
  FROM projects
  WHERE include = 'y'
    AND broad_header = 'Restoration and recovery'
    AND section = 'Recovery and rebuilding programs'
  ORDER BY project_number
")
dbDisconnect(con)

knitr::kable(recovery_projects)
```

### B. Restoration prioritization

```{r restoration-projects}
con <- get_db_connection()
restoration_projects <- dbGetQuery(con, "
  SELECT project_number, source, project_title
  FROM projects
  WHERE include = 'y'
    AND broad_header = 'Restoration and recovery'
    AND section = 'Restoration prioritization'
  ORDER BY project_number
")
dbDisconnect(con)

knitr::kable(restoration_projects)
```

### Template / Guiding Questions

Research Goals

-   What were the key objectives of this research theme?

-   How do these relate to original goals of PSSI/BCSRIF?

-   What major gaps in our knowledge were being addressed?

Outcomes

-   What have we learned and how have we moved the dial on our scientific understanding?

-   What new tools/tech/techniques have been created?

-   How can these projects inform management planning/actions/policy?

Engagement, Collaboration & Knowledge Transfer

-   How did these projects engage with other groups? Was this effective? 

-   Have new ways to ensure knowledge transfer to different audiences been developed?

-   What products have been created? 

Next Steps

-   Major conclusions

-   How could/should the project findings be operationalized for salmon conservation and management?

-   Remaining knowledge gaps and recommendations for future studies