# Summary of research and findings from the Pacific Salmon Strategy Initiative

# Preface {.unnumbered}

```{r, include=FALSE}
# Source common setup (created by build-report.R)
source(here::here("_common.R"))

# Verify data loaded
if (!exists("projects_df")) {
  warning("Project data not loaded. Check _common.R")
  
}
```

## Abstract {.unnumbered}

Brief overview of the report's purpose and findings.

## Citation {.unnumbered}

Lagasse, C., Enns, J., Brown, N., Pearsall, I., Huang, A.-M., and Dobson, D. 2026. Summary of research and preliminary findings from the Pacific Salmon Strategy Initiative. Canadian Technical Report of Fisheries and Aquatic Sciences nnn.

## Report Information {.unnumbered}

**Report Number:** nnn  
**Year:** 2026  
**Region:** Pacific Region  
**DOI:** TBD  
**ISBN:** TBD  
**Catalog Number:** TBD

**Contact Information:**  
Pacific Biological Station  
Fisheries and Oceans Canada  
3190 Hammond Bay Road  
Nanaimo, British Columbia, V9T 6N7, Canada

---
\newpage

## Table of Contents {.unnumbered}
```{r toc, results = "asis"}
# add ToC after the front matter
officer::block_toc(level = 3)
```

\newpage