---
title: "Summary of research and preliminary findings from the Pacific Salmon Strategy Initiative"

subtitle: |
  Canadian Technical Report of Fisheries and Aquatic Sciences nnnn
  
  Fisheries and Oceans Canada  
  Pacific Biological Station  
  Nanaimo, BC

authors:
  - name: "Cory Lagasse"
    affiliation: "Fisheries and Oceans Canada, Institute of Ocean Sciences, Sidney, BC, Canada"
  - name: "Joe Enns"
    affiliation: "Fisheries and Oceans Canada, Pacific Biological Station, Nanaimo, BC, Canada"
  - name: "Nicholas Brown"
    affiliation: "Fisheries and Oceans Canada, Pacific Biological Station, Nanaimo, BC, Canada"
  - name: "Isobel Pearsall"
    affiliation: "Fisheries and Oceans Canada, Pacific Biological Station, Nanaimo, BC, Canada"
  - name: "Ann-Marie Huang"
    affiliation: "Fisheries and Oceans Canada, Pacific Biological Station, Nanaimo, BC, Canada"
  - name: "Diana Dobson (editors)"
    affiliation: "Fisheries and Oceans Canada, Pacific Biological Station, Nanaimo, BC, Canada"

date: "2026"
---


```{r table-settings}
# recommended table settings to match CSAS table style and accessibility standards
flextable::set_flextable_defaults(
  font.size = 8,
  padding.left = 3, 
  padding.right = 3, 
  padding.top = 1, 
  padding.bottom = 1,
  na_str = "-",
  theme_fun = "theme_booktabs"
)
```

`r officedown::knit_print_run(officer::run_pagebreak())`

<!--chapter:end:index.Rmd-->

---
# Development knitting only - ignored by Quarto book build
output:
  word_document:
    reference_docx: templates/custom-reference.docx
bibliography: bib/references.bib
csl: csl/csas.csl
link-citations: true
---

```{r, include=FALSE}
source(here::here("_common.R"))

# Add this line to allow duplicate labels in child documents
options(knitr.duplicate.label = "allow")

```

# Introduction {#sec:introduction}

## Pacific Salmon Strategy Initiative Overview

The Pacific Salmon Strategy Initiative (PSSI) was a generational investment of over \$500 million in habitat restoration, hatchery enhancement, fisheries management, scientific research, and other activities related to Pacific salmon. Over its five year span beginning in 2021, the multi-faceted initiative sought to restore vulnerable populations of Pacific salmon and their habitats, support sustainable fishing opportunities while reducing impacts on at‑risk stocks, and build capacity among partners to achieve better outcomes for Pacific salmon. In doing so, the initiative sought to respond to long‑standing concerns about declines in many salmon populations, increasing pressures from climate change, and the need for more integrated approaches to management and recovery.

PSSI brought significant new capacity and resources to Fisheries and Oceans Canada (DFO), the lead federal agency for the initiative, enabling expanded scientific research, enhanced stewardship and restoration activities, additional capacity for enhancement and monitoring, and new collaborative approaches with Indigenous and non-Indigenous partners. The initiative was initially designed and managed across four pillars—Conservation and Stewardship, Salmon Enhancement, Harvest Transformation, and Integration and Collaboration—each of which involved multiple sectors within DFO and included several initiatives (**Table** \@ref(tab:pssi-initiatives)). The pillars were intended to provide a framework for aligning actions across habitat, hatcheries, fisheries, and governance in support of the long‑term persistence and resilience of Pacific salmon.

The conservation and stewardship pillar focused on new activities and investments supporting the long-term persistence and rebuilding of Pacific salmon populations. Key components included a doubling of the British Columbia Salmon Restoration and Innovation Fund (BCSRIF), investing in research to improve our understanding of salmon and their ecosystems, advancing integrated planning for salmon ecosystems to mitigate pressures on freshwater habitats, developing integrated rebuilding plans for vulnerable salmon populations, improving supports for salmon stewardship activities through engagement, and capacity-building with First Nations and community partners.

The salmon enhancement pillar aimed to modernize hatchery programs to more effectively support the conservation, recovery, and sustainable use of Pacific salmon. A key objective under this pillar was to invest in new hatchery infrastructure and capacity, including retrofitting existing facilities and building new facilities in the Upper Fraser area. The pillar also aimed to strengthen collaboration and enhancement partnerships, including with First Nations and other stakeholders, while investing in hatchery science related to genetics, fish health, salmon ecosystems, and operational monitoring. In addition, this pillar supported development of a modernized enhancement policy intended to guide decision-making in a way that better supports the overall enhancement vision, including the targeted use of conservation-based hatchery production to help recover, stabilize, and restore vulnerable salmon stocks.

The harvest transformation pillar was intended to change how Pacific salmon fisheries were managed in response to long-term stock declines and ongoing conservation concerns. Its key objectives included adopting more precautionary and selective fishing approaches, including exploring mark-selective fisheries, and developing new tools to better manage harvest capacity across sectors. The pillar also included commitments to update the Salmon Allocation Policy and to update licensing, regulatory, and management frameworks, including measures such as licence buyback programs and long-term commercial fishery closures in areas with significant conservation concerns. In addition, it aimed to improve the timeliness and accuracy of fishery monitoring and catch reporting programs to strengthen the information base used for harvest decision-making.

The integration and collaboration pillar focused on strengthening collective capacity to address Pacific salmon declines by improving how DFO worked both internally and with external partners. Key objectives under this pillar were to provide capacity support to various internal and external groups involved in salmon conservation and management, to improve collaboration among federal, provincial, territorial, Indigenous, and other governments, and to prioritize salmon stocks for recovery through coordinated, cross-jurisdictional planning and decision-making processes.

```{r pssi-table, tab.cap="The 24 PSSI Initiatives by pillar.",tab.id="pssi-initiatives",tab.topcaption=TRUE}

pillars_df <- read_csv(here(rawdata_path, "Table1_PSSI_initiatives.csv")) %>%
  mutate(across(where(is.character), ~ str_replace_all(.x, "[^[:alnum:][:punct:][:space:]]", " ")))

ft <- flextable(
  pillars_df,
  col_keys = c("Pillar", "#", "Initiative Name")
) %>%
  merge_v(j = "Pillar") %>%
  align(j = "Pillar", align = "center", part = "body") %>%
  valign(j = "Pillar", valign = "center", part = "body") %>%
  bold(j = "Pillar", bold = TRUE, part = "body") %>%
  bold(i = ~ `Science lead or support` == "Y", bold = TRUE, part = "body") %>%
  theme_vanilla() %>%
  autofit() %>%
  width(j = "Pillar", width = 1.5)


ft

```

## Scientific Research Components of PSSI

The PSSI pillars were underpinned by a dual-track investment in scientific research, combining internal DFO expertise with external grants and partnerships. This new scientific capacity supported a diverse range of projects intended to inform rebuilding plans, identify factors driving population declines, better understand the impacts of climate change, and pilot new approaches for monitoring, enhancement, fisheries harvest, and restoration. By supporting research within DFO and externally, the initiative aimed to leverage and build on existing capacity within the federal government and across academic institutions, Indigenous organizations, non-governmental organizations, and industry.

While the first phase of BCSRIF was already underway with 97 agreements, PSSI significantly expanded this reach by funding a second phase. This resulted in 72 new collaborative agreements (Appendix \@ref(app:BCSRIF)), with many focused on scientific research in areas such as climate change modelling, habitat restoration effectiveness, and monitoring innovation. These investments enabled First Nations, academic researchers, industry, and community organizations to work alongside DFO scientists, strengthening the evidence base for management decisions while incorporating Indigenous knowledge, local observations, and novel technologies into the broader PSSI research portfolio.

Within DFO Science, PSSI resources were primarily concentrated into two initiatives: Improved Understanding of Salmon Ecosystems and Science to Support Hatchery Modernization. Targeted funding also bolstered research in support of the Habitat Restoration Centre of Expertise, improving the management and accessibility of salmon-related data, increasing the capacity of the molecular genetics lab, and expanding monitoring of recreational fisheries. This work was implemented across the Pacific Region’s four Science divisions—Stock Assessment, Ecosystem Science, Ocean Science, and Aquatic Diagnostics—enabling multi-disciplinary approaches to research and monitoring.

The largest PSSI initiative led by DFO Science was Improved Understanding of Salmon Ecosystems. The overarching aim of this initiative was to enable the development of effective conservation strategies and management actions for wild salmon and their ecosystems by developing a comprehensive understanding of habitat, ecosystem, and climate change impacts on salmon productivity. Research planning and prioritization began in 2022 with a DFO workshop that identified knowledge gaps and research themes [@lagasse2024]. Subsequently, research proposals underwent a peer-review process and steering committee evaluation. Beginning in 2023, dozens of multi-year projects were funded, including the establishment of new research programs and the expansion or continuation of existing long-term programs.

The second major DFO Science initiative, Science to Support Hatchery Modernization, was a part of the PSSI Salmon Enhancement pillar. This initiative was designed to provide scientific advice and information to inform planning and program implementation, advance the use of molecular genetics and other tools for assessment, and expand fishery and escapement sampling programs. These research activities were intended to support the expansion of hatchery programs under PSSI, informing effective enhancement for population rebuilding and providing harvest opportunities while reducing risks for wild salmon populations. Research priorities were co-developed by a steering committee representing the Salmonid Enhancement Program and DFO Science.

In this report, we summarize the outcomes and findings of research projects funded through the PSSI, with a focus on work led by DFO Science in the Pacific Region. We synthesize results across several research themes aligned with the major PSSI initiatives(Figure 1), providing an overview of findings from a diverse set of multi-disciplinary projects. For the Improved Understanding of Salmon Ecosystems initiative, we review research within five thematic areas: salmon population monitoring, life-cycle approaches, freshwater ecosystems, marine ecosystems, and climate change vulnerability. For the Science to Support Hatchery Modernization sub-initiative, we outline research conducted under the Genetics, Fish Health, and Hatchery Practices themes. In addition, we describe broader research and monitoring activities supported by PSSI, including work related to fisheries monitoring and new harvest techniques, stock assessment advice, and new tools for data mobilization and management. We provide project summaries for `n_projects` DFO-led research projects, offering a snapshot of the current state of Pacific salmon science under PSSI. Although the synthesis is not comprehensive of all research conducted under PSSI, it highlights the insights, knowledge gaps and future priorities identified during the five years of the initiative as they relate to thematic areas.

\newpage

# Improved Understanding of Salmon Ecosystems {#sec:salmon-ecosystems}

### A. Salmon population monitoring

```{r salmon-population-projects, tab.cap="DFO and BCSRIF projects associated with salmon population monitoring.",tab.id="tbl-salmon-pop"}

make_section_table(projects_df,
                   section_pick = "Salmon population monitoring")


```

### B. Life cycle approaches

#### Follow the Fish (IP)

Provide an overview of the follow the fish program, how risk assessments and limiting factors (RAMS) were used to guide research priorities, key insights, and next steps to support rebuilding of WCVI Chinook salmon.

```{r ftf-projects, tab.cap="DFO and BCSRIF projects associated with Follow the Fish.",tab.id="tbl-ftf"}

make_section_table(projects_df,
                   section_pick = 'Life cycle assessment - Follow The Fish')

```

#### Bottlenecks to survival (IP)

```{r bottleneck-projects, tab.cap="DFO and BCSRIF projects associated with bottlenecks to survival.",tab.id="tbl-bottlenecks"}

make_section_table(projects_df,
                   section_pick = 'Life cycle assessment - Bottlenecks to survival')

```

### C. Freshwater ecosystems (JE)

```{r freshwater-ecosystem-projects, tab.cap="DFO and BCSRIF projects associated with freshwater ecosystems theme.",tab.id="tbl-fweco"}

make_section_table(projects_df,
                   section_pick = 'Freshwater ecosystems')
```

Summarize research undertaken and key findings focused on freshwater ecosystems. Projects and themes include:

-   Risk assessment and mitigation of aquatic invasive species

    -   Understanding threats of freshwater AIS (Thomas Therriault)

    -   Survival and predation on Okanagan Chinook (Tommy Pontbriand)

    -   Monitoring of Smallmouth bass in Cultus Lake (Daniel Doutaz)

-   Research on freshwater toxins and their effects on salmon, particulalrly 6PPD-Q

    -   Identifying tire-related toxic hotspots (Lisa Loseto)

-   Physiological and mechanistic studies to link environmental conditions with survival and productivity

    -   Mechanistic modelling linking hydrology to habitat quality (Sean Naman)

    -   Physiological studies to understand cumulative impacts on salmon survival (David Patterson)

    -   Effects of redd scour on WCVI Chinook (Diana McHugh)

-   Broad-scale spatial and statistical modelling to characterize cumulative threats to freshwater habitat

    -   Geospatial indicators of cumulative threats to stream habitat (Josie Iacarella)

    -   Effects of forest harvest of salmon productivity (BCSRIF, Maria Kuruvilla)

    -   The CEMPRA cumulative effects modelling framework (BCSRIF, Jordan Rosenfeld)

### D. Marine ecosystems (IP/NB)

```{r marine-ecosystem-projects, tab.cap="DFO and BCSRIF projects associated with marine ecosystems and stressors theme.",tab.id="tbl-marine-eco"}

make_section_table(projects_df,
                   section_pick = 'Marine ecosystems')
```

Summarize research and key findings related to building understanding of marine ecosystems and key stressors. Projects and themes include:

-   Oceanographic studies monitoring marine conditions to understand their influence on salmon productivity

    -   Zooplankton and krill monitoring (Akash Sastri)

    -   Hypoxia and temperature within Vancouver Island fjords (Laura Bianucci)

    -   Acoustic monitoring of salmon and salmon prey (Stephane Gauthier)

    -   Monitoring harmful algal biotoxins (Andrew Ross)

    -   Contaminants monitoring (Lisa Loseto)

-   Research and monitoring on marine ecosystems and predators

    -   eDNA monitoring in WCVI (Christoph Deeg)

    -   Quantifying salmon predation by sea lions (Came ron Freshwater)

### E. Climate change vulnerability (CL)

```{r climate-projects, tab.cap="DFO and BCSRIF projects associated with climate change vulnerability.",tab.id="tbl-climate"}

make_section_table(projects_df,
                   section_pick = 'Climate change vulnerabiliity')
```

**Highlights**

-   Newly developed marine and freshwater climate models allow us to assess environmental changes that will be experience by Pacific salmon within British Columbia

-   Vulnerability assessments reveal differences in risk for the 50 conservation units of Pacific salmon within the Fraser basin, with Pink and Chum salmon the least vulnerable to climate changes while ….

-   Life stage models….

-   Genetic adaptation….

-   Greenhouse gas emissions matter. Vulnerability, as represented by conditions beyond optimal thresholds, was more severe in high emissions scenarios compared to scenarios with mitigation.

-   These new assessments can be applied in planning and decision-making to contextualize risks and benefits….

**Research Goals**

The significance of climate change for Pacific salmon, and biodiversity in general, cannot be overstated. Studies across their range have highlighted the vulnerability of Pacific salmon to recent and future climate changes for freshwater and marine life stages [@mantua2010; @mcclure2023; @crozier2019; @grant2019]. Climate change emerged as a key theme during the PSSI Science planning workshop [@lagasse2024] and was a central element of the Improved Understanding of Salmon Ecosystems sub-initiative. Discussion papers, minister mandate letters, funding initiatives, and national adaptation plans have recognized the need to make climate change considerations central to decision-making and planning [@macdonald2023]. This need follows from the understanding that climate change introduces additional risk and uncertainty into plans and decisions because expectations based on past climate relationships may not hold true.

Yet there is still much that is uncertain about how climate changes will unfold for Pacific salmon, such as which species and stocks will be most affected, when physiological thresholds or ecological tipping points will be reached, the relative influence of various environmental changes on productivity, and the degree to which populations will be able to evolve and adapt to changing conditions and increasing temperatures [@schoen2017; @chittenden2009; @crozier2025]. Research within this theme sought to better understand and resolve these uncertainties through a variety of approaches including downscaled climate modelling, modern genotyping, life-cycle modelling, and vulnerability risk assessments. Collectively, these studies reveal important details about the implications of climate change for Pacific salmon that can allow us to understand and anticipate current and future changes.

**Downscaled climate modelling**

Global climate modelling – the simulation of earth systems into the future under different greenhouse gas emission scenarios – is the basis for much of our understanding about future climate change as reported on by the intergovernmental panel on climate change (IPCC). However, these global models lack the necessary detail to tease out fine-scale environmental changes to aquatic habitats that will be significant to understanding responses of Pacific salmon over the coming decades. Recent work, including many projects supported by PSSI, is addressing these gaps through downscaling of global climate models to marine and freshwater environments across the Pacific region.

Within freshwater habitats, the Pacific Climate Impacts Consortium (PCIC) has developed high resolution stream temperature and flow models that project future conditions throughout British Columbia. Statistically downscaled models of August stream temperature have also recently been developed providing an alternative method for projecting future temperatures.  These models reveal a gradient of temperature changes throughout British Columbia influenced not only by global climate patterns, but by elevation, glacial coverage, latitude…..

Within marine areas, regional oceanographic models developed by the Ocean Science division project changes in salinity, temperature, and biogeochemistry for waters across the continental shelf of British Columbia. While these models have recently been applied to understand how climate change will affect habitat availability for groundfish and other marine species, applying them towards the assessment of climate change vulnerability for Pacific salmon is more challenging…  Given the importance of marine survival to the productivity of Pacific salmon, this is an important area for further research….

**Vulnerability assessments**

For the purposes of Pacific salmon management, projecting future environmental conditions is only helpful if we can predict how Pacific salmon will respond to those changes. Climate change vulnerability assessments apply a diverse range of methods to consider how the combined exposure and sensitivity of a species or population determines the risks they face from climate change. While such assessments have been applied to thousands of different species at regional and global scales, including to Pacific salmon in the continental US, it is only recently and with support from PSSI that these studies have been applied to Pacific salmon in Canada.

Two of these studies have been focused on the Fraser Basin, the largest and most diverse river basin for Pacific salmon in Canada with over 50 conservation units. The Pacific Salmon Foundation developed a model to quantify life-stage specific exposure of Pacific salmon conservation units to environmental conditions above optimal thresholds, including high temperature and low flow thresholds. DFO Science has developed a climate vulnerability indicators framework that incorporates several downscaled climate models, demographic information such as conservation status, and genetic diversity information to characterize key mechanisms of vulnerability.

Together, these studies reveal …..

Life stage modelling can also be used for evaluating and disentangling the influence of environmental conditions on salmon productivity retroactively and into the future. Models of sockeye salmon, covering populations from the southern to northern ends of their range, reveal important similarities and differences in responses….

**Genetic adaptation**

**Knowledge gaps and future priorities**

Expanding vulnerability assessments to populations beyond the Fraser basin

Further understanding thresholds and tipping points, and how these differ among species of Pacific salmon

Changes in the frequency of extreme events, such as flooding, marine heat-waves, and droughts, are difficult to accurately model and assess but are important determinants of salmon survival

Incorporating climate vulnerability information into decision-making and planning frameworks, prioritizing actions that can mitigate vulnerability

\newpage

# Science to support hatchery modernization (CL/SEP) {#sec:hatcheries}

Describe the enhancement pillar of PSSI, DFO Science sub-initiatives, and the priority research areas. Include a description of the SEP-SCI steering committee and process used to prioritize and develop research projects (adapt elements from the SEP hatchery science booklet).

### A. Genetics

```{r genetics-projects, tab.cap="DFO and BCSRIF projects associated with genetics.",tab.id="tbl-genetics"}
make_section_table(projects_df,
                   section_pick = 'Genetics')
```

Summarize research and key findings from genetics projects. Research questions (from SEP):

-   What are the differences between hatchery and natural origin fish?

-   Why do these differences occur and what are the mechanisms behind them?

-   What factors affect long-term population fitness?

Projects and themes include:

-   Developing and operationalizing modern genetic methods for stock identification and assessment

    -   Continuation/expansion of PBT programs (Eric Rondeau)

    -   Improving stock delineation in Chum salmon by characterizing genome-wide genetic variation (Anna Tigano)

    -   Rapid broodstock screening for improved genetic management of Pacific salmon (Kyle Wellband)

-   Evaluating genetic differences among hatchery and wild populations, and their implications for fitness

    -   Fitness effects of hatchery enhancement and relative reproductive success (Tim Healy)

    -   Epigenetic variation and inheritance in hatchery and natural origin Chinook and Coho salmon (Tim Healy, Kyle Wellband)

    -   Genetic variation linked to early maturation in male Chinook salmon

### B. Fish health and hatchery practices

```{r health-projects, tab.cap="DFO and BCSRIF projects associated with the fish health theme.",tab.id="tbl-health"}

make_section_table(projects_df,
                   section_pick = 'Fish health and hatchery practices')
```

Summarize research and key findings from fish health projects. Research questions (from SEP):

-   How can we better assess and track fish condition and stress?

-   How do we advance aFish health and hatchery best practicesnd modernize diagnostic and health-management practices?

-   How can new molecular and technological tools help us detect issues early and inform decision-making?

Projects and themes include:

-   New diagnostic and monitoring tools for fish health

    -   Expanding and modernizing the finfish diagnostic lab (Amy Long)

    -   Developing a proactive, modernized approach to ensure optimal health and condition (Karia Kaukinen, Kristi Miller)

    -   Using scales to monitor chronic stress in Chinook and Coho salmon (Stewart Johnson)

    -   Application of nanopore technology for detection and characterization of pathogens (BCSRIF, Ahmed Siah)

-   Evaluating factors determining fish health and disease prevalence

    -   Using broodstock data to understand factors that affect BKD rates (Amy Long)

    -   Factors influencing IHNV in sockeye salmon (Kyle Garver)

    -   Modeling interactions of environmental, biological, and infectious factors at the Quinsam river hatchery (Simon Jones)

-   Effects of rearing practices and diet on hatchery fish survival

    -   Optimizing feeds for hatchery production of Pacific salmon (Erika Eliason)

    -   Predicting Chinook reproductive success based on adult thiamine levels (Erika Eliason)

    -   Causes and consequences of vateritic otoliths in hatchery-reared Coho salmon (BCSRIF, Leigh Gaffney)

\newpage

# Fishery monitoring and harvest techniques (NB/CL/FM) {#sec:fishery-monitoring}

```{r fishery-projects, tab.cap="DFO and BCSRIF projects associated with fisheries monitoring and harvest techniques",tab.id="tbl-harvest"}
make_section_table(projects_df,
                   section_pick = 'Fishery monitoring and harvest techniques')
```

#### Highlights

-   New technologies and collaborative partnerships are improving efficiency and accuracy of salmon fishery monitoring programs

-   Research has identified new methods, opportunities, and best practices for selective fishing that can mitigate impacts on non-target stocks and species

#### Background and objectives

The PSSI harvest transformation pillar focused on how precautionary fishery management can help address persistent declines in numerous British Columbia and Yukon salmon stocks. Various efforts and management measures were embraced to improve selectivity of fisheries across all sectors, including advancement of selective harvest methods, reformed licensing and regulatory approaches, enhanced catch monitoring, and long-term closures of some commercial fisheries. Research projects conducted under the harvest transformation pillar focused on developing, testing, and applying improved catch monitoring and reporting frameworks to address longstanding information gaps, particularly with respect to mixed‑stock fisheries. Pilot projects were also conducted to evaluate selective and mark‑selective fishing methods to better understand their effectiveness in reducing impacts on stocks of concern while maintaining harvest opportunities. Collectively, these projects lay the foundation for a refined approach to managing and regulating Pacific salmon fisheries. Preliminary results and management implications from a selection of PSSI harvest transformation research projects are summarized thematically below.

#### Enhancements to recreational monitoring

British Columbia is a renowned destination for salmon angling. Active recreational fisheries span the entire coastline alongside an extensive network of lodges and guiding operations. Likewise, recreational fishing is prevalent and popular in many of BC’s salmon-bearing watersheds. It is a considerable challenge to monitor and assess the impacts to salmon of such widespread, diffuse fisheries, and critical information gaps exist. As a result, several new recreational fishery restrictions have been introduced in recent years to protect declining stocks. PSSI sub-initiative 18.2 focuses on developing or improving recreational fishery monitoring to better inform management actions.

In one such effort, the Central Coast Indigenous Resource Alliance has developed and implemented a creel survey with expanded coverage across a large swathe of the Central Coast. Researchers designed an improved survey and analytical framework that fills gaps in existing DFO programs and is more robust against data limitations. The survey also includes expansive genetic sampling that will better inform the stock composition of salmon catches, particularly Coho, in the region. Finally, this work is strengthening First Nation leadership in fishery monitoring and paves the way for a more active role moving forward.

-   MSF reference fishery (Erin Rechisky)

-   Creel monitoring capacity improvements (Natalie Fuller)

#### Research on release mortality and best practices for recreational fisheries

The Pacific Salmon Ecology and Conservation Laboratory at the University of British Columbia undertook a 5-year research program to study the mechanisms and impacts of injuries sustained by chinook and coho salmon when released by recreational anglers. The BC Sport Fishing Institute, Pacific Salmon Foundation, Canada’s Ocean Tracking Network, DFO, and Kintama Marine Research were key collaborators on this work.

Results arising from acoustic telemetry tracking, extended holding studies, and physiological assays demonstrate that post‑release mortality varies considerably among catch and release methods and that fish condition, species, size, and environmental conditions are also important factors. In broad terms, chinook salmon released in good physical condition generally exhibit low mortality, whereas individuals exhibiting injuries such as bleeding, scale or fin damage, or eye trauma experience substantially higher mortality. Coho salmon appear less resilient compared to chinook released under similar conditions. In addition, smaller fish suffer higher mortality than larger fish, and warm ocean temperatures during capture and release exacerbate losses. These findings indicate that release mortalities in marine recreational fisheries can be significant but that outcomes depend strongly on the circumstances under which fish are captured and released.

Based on these findings, the Pacific Salmon Ecology and Conservation Laboratory is developing practical, evidence‑based recommendations to minimize injury, physiological stress, and subsequent mortality when releasing salmon. Key recommendations emphasize reducing physical contact and air exposure, avoiding the use of landing nets and on‑board handling, releasing fish at the waterline, and minimizing fight times. Gear recommendations include eschewing in-line flashers as well as large or multiple hooks, especially treble hooks,  all of which increase injury risk and severity. Recommendations for fishery management include minimizing fishing in areas dominated by immature fish or salmon predators, winding down fisheries when surface water temperatures are exceptionally warm, and stopping fishing trips once daily retention limits have been achieved. These best practices provide new options to reduce recreational fishing impacts while maintaining opportunity. Researchers are also developing resources to help integrate release mortality considerations into monitoring and public education programs.

#### Bycatch monitoring in commercial fisheries

-   Salmon bycatch monitoring in the trawl fishery (Cory Lagasse)

#### Selective harvest methods and monitoring to support adaptive management

Several research initiatives focused on improving the timeliness, accuracy, and operational relevance of fishery data to support measures aiming to reduce impacts on stocks of concern. A central theme across projects is the integration of emerging technologies with established fishing practices and community‑based monitoring programs. All projects recognized the need for innovation in both fishing practices and monitoring to better support precautionary management.

DFO fishery management indicated that expanded fishery observer coverage is required on the North Coast. This motivated efforts to advance electronic monitoring for commercial vessels as an alternative or offset to the costly at-sea observer program. A pilot study was conducted by the United Fishermen & Allied Workers Union, Teem Fish Monitoring, and Ecotrust Canada  during a Skeena River gillnet demonstration fishery. The electronic monitoring system accurately captured core information, including fishing effort, retained catch, released catch, and fish condition at release. Species identification, camera placement, and potential for crew members to block the view were important limitations, but these might be diminished through improved camera placement, lighting, and fisher cooperation. Future integration of automated image analysis and artificial intelligence could further improve data processing and cost efficiency, emphasizing the value of continuing to develop electronic monitoring systems for commercial salmon fisheries .

Complementary efforts are exploring selective harvest and monitoring tools in First Nation terminal fisheries. A pilot study led by the A‑Tlegay Fisheries Society is revitalizing traditional salmon trap technologies in the Campbell–Quinsam estuary to enable selective harvest of hatchery salmon while safely releasing wild salmon and steelhead. To support improved First Nation terminal fishery management and monitoring, the Pacific Salmon Foundation is developing deep‑learning approaches to automate salmon counting and species identification from video and sonar data, with the goal of linking these tools to escapement monitoring programs on the North and Central Coast. By enabling near‑real‑time data integration, this work supports more responsive management of terminal fisheries. Together, these projects illustrate how selective fishing methods, combined with modern monitoring technologies and Indigenous leadership, can yield more flexible and precautionary fisheries.

\newpage

# Stock assessment - Fisheries Science Advisory Reports (AMH) {#sec:fsar}

Summarize policy context and rationale for FSARs. Describe Salmon Sprint Weeks that occurred during PSSI and FSARs that were completed or are still underway. Describe emergent themes, trends, or challenges across FSARs and stocks that were assessed.

Include a table or figure with list of FSARs.

\newpage

# Data management and tools {#sec:data-management}

Describe the data-related PSSI sub-initiatives and key projects to improve management and availability of data for Pacific salmon.

### A. Improving data management and availability

```{r innovation-projects, tab.cap="Data tools and management applications developed by the Fisheries and Assessment Data Section as part of PSSI.",tab.id="tbl-data-projects"}

data_projects_loc <- here(rawdata_path, "PSSI_data_projects.csv")

# Load + clean + keep some fields
data_projects_table <- read_csv(data_projects_loc) %>%
  mutate(
    Title  = str_squish(Title),
    Status = str_squish(Status),
    Access = str_squish(Accessibility),
    URL    = URL_link
  ) %>%
  select(Title, Status, Access, URL)

ft <- flextable(data_projects_table) %>%
  set_header_labels(Title="Title", Status="Status", Access="Access", URL="URL") %>%
  bold(part="header") %>%
  compose(j="URL", value = as_paragraph(hyperlink_text(x = "Link", url = URL))) %>%
  width(j="Title",  width=3) %>%
  width(j="Status", width=1.1) %>%
  width(j="Access", width=1.4) %>%
  width(j="URL",    width=0.8)

ft

```

### B. Data and modelling to support decision making

```{r data-projects, tab.cap="DFO and BCSRIF projects associated with data and modelling in support of decision making",tab.id="tbl-decisions"}

make_section_table(projects_df,
                   section_pick = 'Data and modelling to support decision making')

```

\newpage

# Summary

How have PSSI Research elements met its objectives? How has work completed moved the dial for salmon conservation and understanding factors driving salmon declines?

# Acknowledgements

\newpage

<!-- Add acknowledgments here -->

<!--chapter:end:01_body.Rmd-->

---
# Development knitting only - ignored by Quarto book build
output:
  word_document:
    reference_docx: templates/custom-reference.docx
bibliography: bib/references.bib
csl: csl/csas.csl
link-citations: true
---

# References {#sec:references}

::: {#refs custom-style="Bibliography"}
:::

\newpage

<!--chapter:end:99_references.Rmd-->

---
# Development knitting only - ignored by Quarto book build
output:
  word_document:
    reference_docx: templates/custom-reference.docx
bibliography: bib/references.bib
csl: csl/csas.csl
link-citations: true
---

# Project Reports {#sec:project-reports}

```{r, include=FALSE}
source(here::here("_common.R"))
# Add this line to allow duplicate labels in child documents
options(knitr.duplicate.label = "allow")
```

This section provides detailed reports for each PSSI-funded research project. Reports are organized by project ID and include highlights, background, methods, findings, insights, and next steps.

```{r generate-project-reports, results='asis'}

message("Total projects in database: ", nrow(projects_df))

# Filter for DFO Science projects with include='y'
projects_to_render <- projects_df %>%
  mutate(project_id = as.character(project_id)) %>%
  filter(
    source == "DFO Science",
    include %in% c("y", "Y"),
    !is.na(project_id),
    !is.na(project_title)
  ) %>%
  arrange(project_id)

# Diagnostics
projects_with_content <- projects_to_render %>%
  filter(!is.na(highlights) | !is.na(background) | !is.na(methods_findings))

projects_without_content <- projects_to_render %>%
  filter(is.na(highlights) & is.na(background) & is.na(methods_findings))

message("\n=== Rendering Plan ===")
message("DFO Science projects with include='y': ", nrow(projects_to_render))
message("  └─ With extracted content: ", nrow(projects_with_content))
message("  └─ Metadata only (no content): ", nrow(projects_without_content))

if (nrow(projects_to_render) == 0) {

  cat("\n\n**No DFO Science projects found marked for inclusion.**\n\n")
  cat("Please check that:\n")
  cat("- Projects have source='DFO Science'\n")
  cat("- Projects have include='y' or include='Y'\n\n")

} else {

  cat("\n\n")

  successful_renders <- 0
  failed_renders <- 0

  for (i in seq_len(nrow(projects_to_render))) {

    project <- as.list(projects_to_render[i, ])

    has_content <- !is.na(project$highlights) ||
                   !is.na(project$background) ||
                   !is.na(project$methods_findings)

    content_status <- if (has_content) "[Full Content]" else "[Metadata Only]"

    message("\n--- Rendering ", i, "/", nrow(projects_to_render), ": ",
            project$project_id, " ", content_status, " ---")

    tryCatch({

      result <- knitr::knit_child(
        template_path,
        envir = environment(),
        quiet = TRUE
      )

      cat(result)

      # TRUE Word page break using OpenXML
      #cat("\n\n```{=openxml}\n<w:p><w:r><w:br w:type=\"page\"/></w:r></w:p>\n```\n\n")

      # 2. Insert a SAFE page break that won't corrupt the XML
      cat("\n\n")
      officedown::knit_print_run(officer::run_pagebreak())
      cat("\n\n")

      successful_renders <- successful_renders + 1
      message("    ✓ Success")

    }, error = function(e) {

      cat(paste0("\n\n### Project ", project$project_id, ": ", project$project_title, "\n\n"))
      cat("**Error rendering this project report.**\n\n")
      cat("Error: ", conditionMessage(e), "\n\n")

      # Page break after error too
      cat("\n\n")
      officedown::knit_print_run(officer::run_pagebreak())
      cat("\n\n")
      
      failed_renders <- failed_renders + 1
      message("    ✗ ERROR: ", conditionMessage(e))
    })

  } # end for loop

  message("\n=== Report Generation Complete ===")
  message("Successfully rendered: ", successful_renders, " projects")
  if (failed_renders > 0) {
    message("Failed to render: ", failed_renders, " projects")
  }

} # end if/else
```

<!--chapter:end:10_project-reports.Rmd-->

---
# Development knitting only - ignored by Quarto book build
output:
  word_document:
    reference_docx: templates/custom-reference.docx
bibliography: bib/references.bib
csl: csl/csas.csl
link-citations: true
---

# APPENDIX B. Overview of Phase 2 BCSRIF Projects {#app:BCSRIF}

```{r appendix-b-setup, include=FALSE}
library(here)
source(here("_common.R"))

knitr::opts_chunk$set(
  # Tables in appendix
  tab.cap.pre = "Table B-",
  tab.cap.sep = ": ",
  tab.lp      = "tabB:",
  tab.topcaption = TRUE,

  # Figures in appendix
  fig.cap.pre = "Figure B-",
  fig.cap.sep = ": ",
  fig.lp      = "figB:",     # optional but recommended for a clean namespace
  fig.topcaption = TRUE
)



```

```{r bcsrif-table, echo=FALSE, tab.cap="BCSRIF Projects",tab.id="bcsrif-projects",tab.topcaption=TRUE}
# Load and filter BCSRIF projects
csv_path <- here("data", "raw", "report_project_list.csv")

if (file.exists(csv_path)) {

  raw <- read.csv(
    csv_path,
    stringsAsFactors = FALSE,
    na.strings = c("", "NA")
  ) %>%
    filter(source == "BCSRIF", include == "y") %>%
    select(project_id, project_title, recipient, description, collaborators) %>%
    mutate(
      project_id    = str_squish(as.character(project_id)),
      project_title = str_squish(as.character(project_title)),
      recipient     = str_squish(as.character(recipient)),
      collaborators = str_squish(as.character(collaborators)),
      description   = str_squish(as.character(description))
    ) %>%
    # make sure NAs become empty strings so compose() never sees NULL/NA
    mutate(across(everything(), ~ ifelse(is.na(.x), "", .x)))

  if (nrow(raw) > 0) {

    # 2-column table (Project + wide Description)
    tbl <- raw %>%
      transmute(
        Project      = "",                 # placeholder; filled with compose()
        Description  = description,
        project_id,
        project_title,
        recipient,
        collaborators
      )

    # Optional collaborator line (blank when none)
    collab_line <- ifelse(
      tbl$collaborators == "",
      "",
      tbl$collaborators
    )
    
    
sep_body   <- fp_border(color = "darkgrey", width = 0.5)
sep_header <- fp_border(color = "black", width = 0.9)


    ft <- flextable(tbl %>% select(Project, Description)) %>%
      set_caption("BCSRIF Projects") %>%
      set_header_labels(Project = "Project", Description = "Description") %>%

      # ---- Build the Project stub with bold LABEL chunks only ----
      compose(
        j = "Project",
        value = as_paragraph(
          # Line 1: project id (normal, per your request)
          as_chunk(tbl$project_id),
          as_chunk("\n"),

          # Line 2: Title label (bold) + value (normal)
          as_chunk("Title: ", props = fp_text(bold = TRUE)),
          as_chunk(tbl$project_title),
          as_chunk("\n"),

          # Line 3: Recipient label (bold) + value (normal)
          as_chunk("Recipient: ", props = fp_text(bold = TRUE)),
          as_chunk(tbl$recipient),

          # Line 4 (optional): Collaborators label (bold) + value (normal)
          as_chunk(ifelse(collab_line == "", "", "\n")),
          as_chunk(ifelse(collab_line == "", "", "Collaborators: "),
                   props = fp_text(bold = TRUE)),
          as_chunk(ifelse(collab_line == "", "", collab_line))
        )
      ) %>%

      # ---- Styling: readable & report-friendly (portrait Word, normal margins) ----
      bold(part = "header") %>%
      align(part = "header", align = "left") %>%

      fontsize(part = "all", size = 9) %>%
      line_spacing(part = "all", space = 1.0) %>%
      padding(part = "all", padding = 3) %>%
      align(part = "all", align = "left") %>%
      valign(part = "all", valign = "top") %>%
      
      border_inner_h(border = sep_body, part = "body") %>%   
      hline(i = 1, border = sep_header, part = "header")  %>%  

      # Stable Word rendering + wide Description
      width(j = "Project",     width = 2) %>%
      width(j = "Description", width = 4.5)

    ft

  } else {
    cat("*No BCSRIF projects found with include = 'y'.*\n")
  }

} else {
  cat("*Project data file not found at:* ", csv_path, "\n")
}

```

\newpage

<!--chapter:end:12_appendix-b.Rmd-->


# APPENDIX C. Summary of DFO Science PSSI Data Projects {#app:data}

```{r appendix-c-setup, include=FALSE}
source(here::here("_common.R"))


knitr::opts_chunk$set(
  # Tables in appendix
  tab.cap.pre = "Table C-",
  tab.cap.sep = ": ",
  tab.lp      = "tabC:",
  tab.topcaption = TRUE,

  # Figures in appendix
  fig.cap.pre = "Figure C-",
  fig.cap.sep = ": ",
  fig.lp      = "figC:",     # optional but recommended for a clean namespace
  fig.topcaption = TRUE
)



# Helper function for including figures from data_tools_figures directory
include_data_figure <- function(filename, caption = "") {
  data_fig_dir <- here("figures", "data_tools_figures")
  fig_path <- file.path(data_fig_dir, filename)
  
  if (dir.exists(data_fig_dir) && file.exists(fig_path)) {
    knitr::include_graphics(fig_path)
  } else {
    warning("Figure not found: ", fig_path)
    cat("*[Figure not available: ", filename, "]*\n", sep = "")
  }
}
```

#### 1. Salmon Tracking, Escapement, Assessment and Management Platform

The Salmon Tracking, Escapement, Assessment and Management ([STREAM](https://pac-salmon.dfo-mpo.gc.ca:8443/bcsn/#/)) Platform is a new online resource that helps improve how salmon escapement information is collected, shared, and understood. [STREAM](https://pac-salmon.dfo-mpo.gc.ca:8443/bcsn/#/) brings together data, tools, and easy‑to‑use applications to support the conservation and management of Pacific salmon.

Fisheries and Oceans staff require timely, consistent access to salmon data across sectors. The STREAM platform enhances the timeliness of salmon count data, standardizes the spatial organization of salmon information, and improves reporting on salmon status. This foundational information supports stock assessment biologists, fisheries managers, species at risk biologists, and other decision-makers in delivering informed, data-driven management.

The platform has three main areas: entering and uploading data, learning about salmon populations, and exploring status reports (Figure 1). Through these features, users can view information about salmon populations across different regions and access data that support science, monitoring, and decision‑making.

As STREAM continues to grow, additional funding would allow us to respond to user feedback, develop more tools that improve data quality, and make more information available to the public.

```{r stream-figure, echo=FALSE, out.width=NULL, out.height=NULL, fig.cap="Screen capture of the Salmon Tracking Escapement Assessment and Management (STREAM) Platform with the table of contents visible."}

include_data_figure("Figure1.png")
```

#### 2. Salmon Space

[Salmon Space](https://egispi.ent.dfo-mpo.ca/apps/SalmonSpace-Espacesaumon/) is an interactive map that lets you explore salmon information across British Columbia and Yukon. You can search by location or by salmon species to find data such as salmon counts and stock status (Figure 2).

There is strong public demand for access to salmon data to enhance transparency and build trust in departmental decision-making. Salmon Space provides up-to-date salmon counts and status information in an accessible format. By improving data transparency, this application strengthens trust and alignment between internal and external biologists, Indigenous partners, and the broader community of salmon stakeholders and enthusiasts.

The map includes several optional layers, including census sites, conservation units, stock management units, and designatable units, allowing users to view salmon information at different scales. Data is updated weekly from the Fisheries and Oceans regional database, so the public can access recent information with confidence.

Users can download both tabular and spatial data directly from the site, or customize and print map views to support reports, presentations, and research.

```{r salmon-space-figure, echo=FALSE, out.width=NULL, out.height=NULL, fig.cap="Screen capture of Salmon Space showing the Osoyoos-Skaha-Okanagan Sockeye conservation unit (highlighted) with associated information displayed in a pop-up."}
include_data_figure("Figure2.png")
```

#### 3. Specimen Tracking and Analysis Management Platform (STAMP)

The Specimen Tracking and Analysis Management Platform (STAMP) is an integration and tracking system that supports field biologists, laboratories, and analysts by linking information about individual organism specimens across field sampling, tissue handling, and laboratory analysis. STAMP does not collect new data; it connects existing data systems like CREST, KREST and FOS with results data systems like Otomanager and the newly developed Genetics database to ensure that specimen provenance, measurements, and lab results can be reliably tracked and integrated.

Salmon stock assessments and other biological analyses are currently limited by fragmented data systems, inconsistent identifiers, and manual workflows that make it difficult to track tissue samples and link lab results back to individual organisms. STAMP addresses these challenges by providing a standardized unique specimen identifier and shared tracking infrastructure, improving data quality, traceability, and interoperability across projects and systems.

STAMP consists of an integration layer that connects field data systems, laboratory workflows, and results databases, along with tools that support specimen tracking, status reporting, and data linkage. These components enable users to follow specimens from field collection through laboratory processing and to join lab results with organism-level and project-level information.

As STAMP evolves, future development will focus on expanding support to additional lab types, species, and data systems, strengthening data standards and interoperability, and improving tools for querying, reporting, and stewardship. This will enable more efficient operations, reduce manual effort, and support the integrated analyses required by DFO's modern mandate. 

```{r stamp-figure, echo=FALSE, out.width=NULL, out.height=NULL, fig.cap="Screen capture of STAMP."}
include_data_figure("Figure3.png")
```

#### 4. SILscanner: Stream Inspection Log Digitization Platform

SILscanner is a cloud-based digitization and data recovery platform focused on converting paper Stream Inspection Logs (SILs) into machine-readable data. Stream Inspection Logs or SILs are records of how many salmon of each species were observed and counted during a sampling event on a particular day. SILs remain a significant component of salmon monitoring and assessment operations, and SILscanner supports operational programs by transforming these paper records into accessible digital data.

A large volume of salmon monitoring data continues to be captured on paper field forms, which are later keypunched by DFO staff. This creates delays in data availability, limits data quality control, and requires substantial manual transcription effort. SILscanner addresses this challenge by providing a standardized optical character recognition and form-based data extraction pipeline that recovers structured data directly from Stream Inspection Logs, reducing manual data entry and the risk of transcription errors.

The SILscanner platform consists of a cloud-based optical character recognition (OCR) and form-processing pipeline, secure cloud storage for digitized documents, and a database for indexing extracted data. The system is optimized for form-based data capture from SILs, while also supporting the digitization of handwritten and typewritten archival documents and data extraction from text-based PDFs. Digitized records are indexed and made accessible to analytical and text-mining tools, enabling downstream analysis and integration with other salmon data systems.

SILscanner has been rolled out in the Fraser, West Coast Vancouver Island (WCVI), and North Coast (NC) areas and continuing development to scale to additional areas and form types. Future development will focus on expanding regional coverage, improving data extraction accuracy and validation workflows, and strengthening integration with downstream databases. These enhancements will accelerate data availability, improve data quality, and modernize a critical but currently paper-based component of salmon data capture.

```{r silscanner-figure, echo=FALSE, out.width=NULL, out.height=NULL, fig.cap="Screen capture of SILScanner."}
include_data_figure("Figure4.png")
```

#### 5. DocFlow: AI-Assisted Document Processing for Regulatory and Program Data

DocFlow is an AI-assisted document processing platform designed to transform large collections of unstructured PDF documents into searchable, analyzable, and decision-ready data. It supports programs such as Fish and Fish Habitat Protection by unlocking information embedded in regulatory, monitoring, and authorization documents that are currently difficult and time-consuming to use.

Many critical program datasets, including those stored in systems like Program Activity Tracking for Habitat (PATH), exist primarily as unstructured PDF files with limited metadata. Retrieving information such as project impacts, habitat footprints, mitigation measures, or species at risk often requires manual review of large numbers of documents, taking weeks and producing inconsistent results. DocFlow addresses this challenge by combining optical character recognition, artificial intelligence, and subject-matter-expert guidance to extract structured data from documents in a transparent and reproducible way.

The DocFlow platform consists of a document ingestion and OCR pipeline, AI-based information extraction tools, and an expert-guided interface that allows users to define, validate, and refine how key information is identified within documents. Extracted data are stored in structured formats and linked across related documents using graph-based representations, enabling efficient search, analysis, and structuring across large document collections.

Initially developed as a proof of concept for PATH authorization documents, DocFlow is being scaled to additional document types, programs, and regions, supporting integrative analyses such as cumulative effects assessment, regulatory performance reporting, and long-term data stewardship.

```{r docflow-figure, echo=FALSE, out.width=NULL, out.height=NULL, fig.cap="Screen capture of DocFlow."}

include_data_figure("Figure5.png")
```

#### 6. DFO Salmon Data Standards

The Government of Canada (GC) Fisheries and Oceans Canada (DFO) Salmon Ontology is a W3C-published semantic schema that provides a common, machine-readable framework for describing Pacific salmon data across DFO. It defines core concepts—such as stocks, surveys, samples, measurements, methods, and results—and their relationships using a hybrid Web Ontology Language (OWL) and Simple Knowledge Organization System (SKOS) approach. The ontology is schema-only by design, meaning it contains no data values, which supports clean versioning, reuse across systems, and interoperability with external platforms.

The ontology's purpose is to reduce friction in data integration, discovery, and reuse by providing scientists, data stewards, and managers with a shared, well-defined vocabulary. It aligns with international standards including Darwin Core, Basic Formal Ontology (BFO), Information Artifact Ontology (IAO), Sensor, Observation, Sample, and Actuator (SOSA), Provenance Ontology (PROV), and the InteroperAble Descriptions of Observable Property (I-ADOPT) framework. This alignment supports FAIR (Findable, Accessible, Interoperable, Reusable) and CARE principles while enabling future automation such as validation, provenance tracking, and semantic querying without locking DFO into a single system or tool.

#### 7. Salmon Population Summary Database (SPSR)

The Salmon Population Summary Database (SPSR) is a regional repository designed to centralize key salmon population index data—such as spawner abundance, catch, recruitment, exploitation rates, age structure, and hatchery contribution—that underpin routine stock assessments and forecasting. It addresses a major program gap where derived index data are currently scattered across local spreadsheets and bespoke systems, limiting transparency, efficiency, and trust. SPSR is intended to support Fisheries Science Advisory Reports (FSARs) and related assessment work by providing a consistent, FAIR-aligned source of core indices and associated metadata.

The database compiles standardized population indices and reference points at Conservation Unit (CU) and Stock Management Unit (SMU) scales, without replacing authoritative source systems such as NuSEDS, Fishery Operations System (FOS), or coded-wire tag (CWT) databases. It documents methods, assumptions, infilling approaches, and data quality to support reproducibility, succession, and training. SPSR is modeled on established repositories used by the National Oceanic and Atmospheric Administration (NOAA) and is intended for DFO science and assessment staff, with future development focused on expanding coverage, improving interoperability, and supporting strategic reviews of monitoring and assessment frameworks.

#### 8. Open Science Documentation Hub

The Data Stewardship Unit (DSU) is part of the Fishery and Assessment Data Section (FADS) within the Pacific Region Science Branch of Fisheries and Oceans Canada (DFO). The Open Science Documentation Hub provides a public, authoritative reference for data stewardship practices that support salmon science and assessment. It addresses common challenges related to inconsistent data management, limited documentation, and barriers to reproducible analysis across programs.

The hub supports biologists, analysts, and data stewards by documenting practical guidance, standards, and workflows across the full data lifecycle—from planning and collection to analysis and publication. Content is grounded in real biological and management contexts, including stock assessment and monitoring, and is aligned with FAIR data principles and Indigenous data governance considerations such as OCAP®. The site complements internal FADS resources by making core guidance openly accessible.

#### 9. FADS Salmon Data Wiki

The FADS Salmon Data Wiki is an internal, community-maintained reference for DFO Pacific Region Science staff that documents the salmon data landscape across programs, databases, and systems. Its purpose is to support orientation, onboarding, and day-to-day work by providing clear, factual descriptions of datasets, infrastructure, and contacts, without replacing authoritative databases or formal scientific documentation.

The wiki organizes information by major data domains (e.g., fisheries, escapement, genetics, enhancement, habitat) and describes how key systems are accessed and used in biological and management contexts such as stock assessment and fisheries management. It is available to all DFO staff and relies on ongoing contributions to remain current. Future work includes improving consistency across entries, strengthening links to authoritative metadata and open data records, and expanding coverage as systems evolve.

#### 10. Custom Salmon Data GPT

This supports salmon biologists and data stewards by converting messy spreadsheets into standardized Salmon Data Packages (SDPs) that follow a common specification and shared ontology. It streamlines biological and management data workflows by generating consistent metadata, controlled vocabularies, and semantically enriched variable definitions aligned with the DFO Salmon Ontology, Darwin Core, and I-ADOPT measurement patterns. This approach modernizes salmon data management by embedding ontology-based methods and reproducible standards directly into day-to-day data preparation, improving comparability across stock assessment, monitoring, genetics, and fisheries management datasets.

The project is accessed as a Custom GPT in ChatGPT and is intended for DFO scientists, data stewards, and collaborators, with outputs that are openly reusable as CSV-based SDPs. Next steps focus on expanding ontology coverage, refining measurement patterns, and further automating validation through companion tools (e.g., metasalmon). Simple visuals—such as SDP workflow diagrams or ontology relationship graphics—are often included to help users quickly understand how raw data are transformed into standardized, reusable packages.

#### 11. Salmon Data Package (SDP)

The Salmon Data Package (SDP) is a lightweight, frictionless-data-style specification for exchanging salmon datasets between scientists, assessment biologists, and data stewards. It is designed to work with familiar formats such as Excel and CSV while remaining ontology-aware, linking columns and code lists to the DFO Salmon Ontology and related vocabularies. SDP provides a consistent way to describe datasets, tables, variables, and categorical codes without enforcing a single rigid schema.

SDP enables automated validation, reproducible analysis, and interoperability across projects by standardizing metadata and semantics. It supports AI-assisted workflows, integration with R and Python tools (including the metasalmon package), and alignment with standards such as Frictionless Data and Darwin Core. The specification is openly available, versioned, and intended for use by DFO staff and collaborators preparing data for analysis, sharing, publication, or future knowledge-graph integration.

#### 12. Salmon Escapement Estimates Classification Toolkit

The Salmon Escapement Estimates Classification Toolkit is an interactive R Shiny application designed to standardize how salmon escapement estimate types are classified across DFO Stock Assessment groups. It responds to long-standing inconsistencies in how estimate types (Type 1–6) have been applied in NuSEDS, which can affect interpretation of data quality and downstream uses such as Wild Salmon Policy (WSP) rapid status assessments. The toolkit operationalizes updated guidance into a transparent, repeatable decision process.

The application guides users through a three-phase, dichotomous classification key that evaluates enumeration methods, estimation methods, and documentation and accuracy requirements. It provides traceable results, clear downgrade logic, and reference visuals to support consistent application across regions and programs. The tool is publicly accessible via ShinyApps.io for DFO staff and collaborators and is supported by structured YAML logic and reproducible documentation outputs. Future work includes refining guidance based on user feedback, aligning outputs more tightly with NuSEDS tables, and supporting broader integration with assessment workflows and training materials.

#### 13. Salmon Outlook Report Automation

The Salmon Outlook Report project modernizes the annual Salmon Outlook process, which produces categorical and numeric forecasts of salmon abundance by Stock Management Unit and Conservation Unit to support harvest planning. It addresses tight timelines and fragmented workflows by streamlining how Outlook data are collected, processed, and transformed into tables, reports, and presentations used by science and management.

The project uses Survey123 for structured data collection and R-based workflows to automate report and presentation generation. It supports semi-automated slide decks tailored to different audiences and produces technical reports using R Markdown and the csasdown format. The tools are intended for DFO science and assessment staff and are actively under development, with next steps focused on refining data inputs, stabilizing scripts, and improving integration with evolving crosswalks and Outlook processes.

#### 14. Genetics Results Database (GRD)

The Genetics Results Database (GRD) is a governed, centralized data product intended to serve as the authoritative source for Genetic Stock Identification (GSI) and Parentage-Based Tagging (PBT) results across DFO programs. It addresses fragmentation, inconsistent formats, and version conflicts that currently slow reuse and reduce confidence in genetics outputs. The GRD shifts long-term stewardship responsibility away from analytical labs toward a sustainable, FAIR-aligned data product with clear ownership and accountability.

The GRD supports science, assessment, and management by providing traceable, versioned genetics results with standardized metadata, aligned with FAIR and CARE principles and Indigenous data governance. Governance roles are clearly defined across the data lifecycle, with integration planned across regional and enterprise systems such as STAMP, CREST, and FOS. The database is intended for DFO staff, collaborators, and partners, with phased development beginning in 2025 and broader system integration in 2026. Success is measured by adoption, reduced integration effort, and sustained governance.

\newpage

<!--chapter:end:13_appendix-c.Rmd-->

